---
title: "FinalProject"
author: "Nour El Hamidi, Jonàs Salat Torres"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exploratory data analysis and transformation into stationarity

```{r}
serie=ts((read.table("eurodol.dat")),start=1995,freq=12)
```

```{r}
plot(serie)
abline(v=1995:2020,col=4,lty=3)
```
There are trends, as the mean is not constant. This is not stationary.

### Is variance constant?

```{r}
boxplot(serie~floor(time(serie)))
```
All boxes should have the same height for variance to be constant, so it's not.

```{r}
m<-apply(matrix(serie,nrow=12),2,mean)
v<-apply(matrix(serie,nrow=12),2,var)
plot(v~m)
abline(lm(v~m),col=2,lty=3)
text(m, v, 1995:2020)
```
There is also an upwards trend for the mean variance plot. The variance is higher for higher values of the mean, so we apply log transformation.

```{r}
lnserie=log(serie)
monthplot(lnserie)
```
Monthplot indicates seasonality.
```{r}
m<-apply(matrix(lnserie,nrow=12),2,mean)
v<-apply(matrix(lnserie,nrow=12),2,var)
plot(v~m)
abline(lm(v~m),col=2,lty=3)
text(m, v, 1995:2020)
```
The line is now almost flat.

```{r}
ts.plot(matrix(lnserie,nrow=12),col=1:8)
```
This plot doesn't indicate seasonality but since we saw there is with the monthplot, we will apply a differentiation of 12.
### Is the mean constant?

There is no linear trend, but the mean is still not constant. There are long periods where the values stay above or below a level.
```{r}
d12lnserie=diff(lnserie, lag=12)
plot(d12lnserie)
abline(h=0)
abline(h=mean(d12lnserie),col=2)
```

```{r}
monthplot(d12lnserie)
```
```{r}
ts.plot(matrix(d12lnserie,nrow=12),col=1:8)
```

There is no seasonality now.


```{r}
mean(d12lnserie)
```
The mean is almost 0. To check if it's constant, we will keep applying differentiations until the variance increases and keep the previous iteration.

```{r}
d1d12lnserie=diff(d12lnserie)
plot(d1d12lnserie)
abline(h=0)
abline(h=mean(d1d12lnserie),col=2)
```
```{r}
var(lnserie)
```
```{r}
var(d12lnserie)
```
```{r}
var(d1d12lnserie)
```
The variance is still decreasing.

```{r}
d1d1d12lnserie=diff(d1d12lnserie)
plot(d1d1d12lnserie)
abline(h=0)
abline(h=mean(d1d1d12lnserie),col=2)
```
```{r}
var(d1d1d12lnserie)
```

We can see that the variance increases so we will keep with d1d12lnserie as our time series.

### Plot ACF to check stationarity
```{r}
par(mar = c(5, 4, 4, 2))
acf(d1d12lnserie,ylim=c(-1,1),lag.max=60,col=c(2,rep(1,11)), main ="ACF(serie)")
```
We can see that the ACF decreases fast towards 0, so it depends only on the lag. The (1-B)(1-B^12)log(eurodol) is already stationary! It has constant mean and variance.

## ACF and PACF Analysis

```{r}
par(mfrow=c(1,2))
acf(d1d12lnserie,ylim=c(-1,1),lag.max=60,col=c(2,rep(1,11)), main ="ACF(serie)",lwd=2)
pacf(d1d12lnserie,ylim=c(-1,1),lag.max=60,col=c(rep(1,11),2), main ="PACF(serie)",lwd=2)
par(mfrow=c(1,2))
```
We observe a PACF with a decreasing pattern of exponential form with seasonality 12. While ACF only has one lag different from 0.This behaviour suggests a seasonal MA(1) with s=12. P=1, Q=0
On the regular part, We se that the ACF only has one lag different from 0 (the first) before the seasonality repeats. The PACF has also only the first value clearly different from 0, but the second one could be considered. 
Meaning that we could have the models: 
ARMA(1,0)(0,1)
ARMA(0,1)(0,1)


## Model estimation
